---
title: "Comprehensive Statistical Analysis of Biological Model Data"
author: "Data Analyst"
date: "`r Sys.Date()`"
output: 
  html_notebook:
    toc: true
    toc_depth: 3
    theme: readable
---

# Introduction and Analytical Approach

## Research Context and Analytical Framework

Biological research often requires sophisticated statistical analysis to understand complex relationships between treatments and measured variables. In this analysis, we focus on examining the axis minor length across of cross-sectioned vessels of different treatments and biological models, employing a comprehensive, multi-step statistical approach.

### Analytical Workflow

Our analysis follows a rigorous, systematic approach designed to:

1\. **Characterize Data Distribution**: Understand the underlying statistical properties

2\. **Assess Normality**: Determine appropriate statistical testing methods

3\. **Conduct Comparative Analysis**: Identify statistically significant differences

4\. **Quantify Treatment Effects**: Measure the magnitude of observed variations

### Key Methodological Considerations

-   **Data Heterogeneity**: Recognize potential variations across different biological models
-   **Statistical Robustness**: Use multiple testing approaches to ensure reliable conclusions
-   **Comprehensive Reporting**: Provide transparent, reproducible analysis

# Libraries and Computational Environment

## Rationale for Selected Libraries

### Statistical and Data Manipulation Libraries

-   **ggplot2**: Advanced data visualization
    -   Enables creation of complex, publication-quality graphics
    -   Provides flexible grammar of graphics approach
    -   Critical for representing statistical distributions and comparisons
-   **dplyr**: Data manipulation and transformation
    -   Offers intuitive data filtering and summarization
    -   Enables efficient group-based calculations
    -   Supports clean, readable data processing workflows

### Advanced Statistical Analysis Libraries

-   **FSA (Fisheries Stock Assessment)**:
    -   Provides specialized statistical tests
    -   Includes Dunn's test for post-hoc multiple comparisons
    -   Particularly useful for non-parametric analysis
-   **effsize**:
    -   Calculates effect size measures
    -   Goes beyond p-values to quantify meaningful differences
    -   Provides Cohen's d and Cliff's delta for parametric and non-parametric data
-   **nortest**:
    -   Advanced normality testing
    -   Offers robust methods for assessing data distribution
    -   Includes Anderson-Darling test for large sample sizes

## Computational Environment Setup

```{r setup, include=TRUE}
# Load essential statistical libraries
library(ggplot2)    # Advanced data visualization
library(dplyr)      # Data manipulation
library(FSA)        # Advanced statistical tests (Dunn's test)
library(effsize)    # Effect size calculations
library(nortest)    # Advanced normality testing

# Set global options for consistent reporting
knitr::opts_chunk$set(
  echo = TRUE,      # Show code
  warning = FALSE,  # Suppress warnings
  message = FALSE,  # Suppress messages
  fig.width = 10,   # Default figure width
  fig.height = 6    # Default figure height
)
```

# Data Preparation and Initial Exploration

## Normality Testing Strategy

**Normality Tests Overview:**

-   **Shapiro-Wilk Test**:

    -   Purpose: Assess whether a sample comes from a normally distributed population

    -   Best for: Smaller sample sizes (n \< 5000)

    -   Interpretation:

        -   p-value \> 0.05: Data likely normally distributed

        -   p-value ≤ 0.05: Significant deviation from normal distribution

    -   Strengths: Sensitive to deviations from normality, widely used standard method

-   **Anderson-Darling Test**:

    -   Purpose: Assess normality for larger datasets
    -   Best for: Larger sample sizes (n \> 5000)
    -   Characteristics:
        -   More robust for large datasets
        -   Provides sensitive assessment of tail behavior
        -   Gives more weight to tail regions of distribution

The following function implements a smart normality testing strategy:

```{r normality-function}
# Advanced normality testing strategy
test_normality <- function(x) {
  # Large sample normality test (Anderson-Darling)
  if(length(x) > 5000) {
    ad_test <- ad.test(x)
    return(list(
      test = "Anderson-Darling (n > 5000)",
      statistic = ad_test$statistic,
      p.value = ad_test$p.value
    ))
  } 
  # Standard normality test for smaller samples
  else {
    sw_test <- shapiro.test(x)
    return(list(
      test = "Shapiro-Wilk",
      statistic = sw_test$statistic,
      p.value = sw_test$p.value
    ))
  }
}
```

## Data Import and Exploratory Analysis

```{r data-preparation}
# Read data with careful import
my_data <- read.csv("db.csv", stringsAsFactors = FALSE)

# Create output directories
if(!dir.exists("plots")) dir.create("plots")
if(!dir.exists("results")) dir.create("results")

# Identify unique models and treatments
models <- unique(my_data$model)
treatments <- unique(my_data$treatment)

# Initial data overview
cat("Data Summary:\n")
cat("Total Observations:", nrow(my_data), "\n")
cat("Unique Models:", paste(models, collapse = ", "), "\n")
cat("Treatments:", paste(treatments, collapse = ", "), "\n\n")
```

# Statistical Analysis Loop

## Comprehensive Analysis Strategy

**Comparative Statistical Tests:**

-   **Parametric Tests (for normally distributed data)**:

    -   **One-way ANOVA**:

        -   Purpose: Compare means across multiple groups

        -   Used when: Data is normally distributed with multiple treatment groups

        -   Determines statistically significant differences between group means

        -   Follow-up: **Tukey HSD post-hoc** test if ANOVA is significant

    -   **Independent t-test**:

        -   Purpose: Compare means between two independent groups

        -   Used when: Data is normally distributed with two treatment groups

        -   Assesses whether the means of two groups are significantly different\

-   **Non-Parametric Tests (for non-normally distributed data)**:

    -   **Kruskal-Wallis Test**:

        -   Purpose: Non-parametric alternative to one-way ANOVA
        -   Used when: Data is not normally distributed with multiple treatment groups
        -   Compares median ranks across groups
        -   Follow-up: **Dunn's** **test** for pairwise comparisons if significant

    -   **Wilcoxon Rank Sum Test (Mann-Whitney U Test)**:

        -   Purpose: Non-parametric alternative to independent t-test
        -   Used when: Data is not normally distributed with two treatment groups
        -   Compares median ranks between two independent groups

```{r analysis-loop}
# Open results file for comprehensive reporting
results_file <- file("results/statistical_results.txt", "w")
cat("Comprehensive Statistical Analysis Report\n", file = results_file)
cat("Generated on: ", as.character(Sys.time()), "\n\n", file = results_file)

# Systematic analysis for each model
for (current_model in models) {
  # Filter data for current model
  model_data <- my_data %>% filter(model == current_model)
  
  # Create informative visualization
  p <- ggplot(model_data, aes(x = axis_minor_length, color = treatment)) +
    geom_freqpoly(bins = 30, linewidth = 1) +  # Distribution polygon
    geom_boxplot(
      aes(y = 0, fill = treatment), 
      width = 2, 
      alpha = 0.3, 
      position = position_dodge(width = 0.5)
    ) +
    stat_summary(
      aes(y = 0), 
      fun.data = "mean_cl_normal", 
      geom = "pointrange", 
      color = "black", 
      linewidth = 0.5,
      na.rm = TRUE
    ) +
    labs(
      title = paste("Axis Minor Length Distribution -", current_model),
      subtitle = "Frequency Polygon with Treatment Comparison",
      x = "Axis Minor Length",
      y = "Frequency",
      color = "Treatment",
      fill = "Treatment"
    ) +
    theme_minimal() +
    theme(plot.title = element_text(hjust = 0.5))
  
  # Save high-quality plot
  ggsave(
    filename = paste0("plots/model_", current_model, ".png"),
    plot = p,
    width = 10,
    height = 6,
    dpi = 300
  )
  
  # Descriptive and Inferential Statistics
  treatments <- unique(model_data$treatment)
  num_treatments <- length(treatments)
  
  # Detailed header for each model
  cat(
    "\n\n===== Statistical Analysis: ", current_model, 
    " =====\n", file = results_file, append = TRUE
  )
  
  # Descriptive Statistics
  desc_stats <- model_data %>%
    group_by(treatment) %>%
    summarise(
      n = n(),
      mean = mean(axis_minor_length, na.rm = TRUE),
      median = median(axis_minor_length, na.rm = TRUE),
      sd = sd(axis_minor_length, na.rm = TRUE),
      min = min(axis_minor_length, na.rm = TRUE),
      max = max(axis_minor_length, na.rm = TRUE)
    )
  
  # Write descriptive statistics
  cat("Descriptive Statistics:\n", file = results_file, append = TRUE)
  capture.output(print(desc_stats), file = results_file, append = TRUE)
  
  # Normality and Variance Assessment
  norm_results <- by(
    model_data$axis_minor_length, 
    model_data$treatment, 
    test_normality
  )
  
  cat("\nNormality Assessment:\n", file = results_file, append = TRUE)
  capture.output(print(norm_results), file = results_file, append = TRUE)
  
  # Determine distribution characteristics
  all_normal <- all(sapply(norm_results, function(x) x$p.value > 0.05))
  
  # Statistical Test Selection
  if(num_treatments > 1) {
    if(num_treatments > 2) {
      # Multiple group comparison
      if(all_normal) {
        # Parametric: ANOVA
        anova_result <- aov(axis_minor_length ~ treatment, data = model_data)
        cat("\nANOVA Results:\n", file = results_file, append = TRUE)
        capture.output(summary(anova_result), file = results_file, append = TRUE)
        
        # Post-hoc Tukey if significant
        if(summary(anova_result)[[1]]$`Pr(>F)`[1] < 0.05) {
          tukey_result <- TukeyHSD(anova_result)
          cat("\nTukey HSD Post-hoc Test:\n", file = results_file, append = TRUE)
          capture.output(print(tukey_result), file = results_file, append = TRUE)
        }
      } else {
        # Non-parametric: Kruskal-Wallis
        kruskal_result <- kruskal.test(axis_minor_length ~ treatment, data = model_data)
        cat("\nKruskal-Wallis Test Results:\n", file = results_file, append = TRUE)
        capture.output(print(kruskal_result), file = results_file, append = TRUE)
        
        # Dunn's test for pairwise comparisons
        if(kruskal_result$p.value < 0.05) {
          dunn_result <- dunnTest(axis_minor_length ~ treatment, 
                                  data = model_data, 
                                  method = "bh")
          cat("\nDunn's Post-hoc Test:\n", file = results_file, append = TRUE)
          capture.output(print(dunn_result), file = results_file, append = TRUE)
        }
      }
    } else {
      # Two-group comparison
      group1 <- model_data$axis_minor_length[model_data$treatment == treatments[1]]
      group2 <- model_data$axis_minor_length[model_data$treatment == treatments[2]]
      
      if(all_normal) {
        # Parametric: t-test
        t_test_result <- t.test(group1, group2)
        cat("\nt-test Results:\n", file = results_file, append = TRUE)
        capture.output(print(t_test_result), file = results_file, append = TRUE)
      } else {
        # Non-parametric: Wilcoxon test
        wilcox_result <- wilcox.test(group1, group2)
        cat("\nWilcoxon Rank Sum Test Results:\n", file = results_file, append = TRUE)
        capture.output(print(wilcox_result), file = results_file, append = TRUE)
      }
    }
    
    # Effect Size Calculations
    cat("\nEffect Sizes:\n", file = results_file, append = TRUE)
    treatment_pairs <- combn(treatments, 2, simplify = FALSE)
    
    for(pair in treatment_pairs) {
      g1 <- model_data$axis_minor_length[model_data$treatment == pair[1]]
      g2 <- model_data$axis_minor_length[model_data$treatment == pair[2]]
      
      # Choose appropriate effect size method
      if(all_normal) {
        es <- cohen.d(g1, g2)
        cat(paste("\nCohen's d for", pair[1], "vs", pair[2], ":\n"), 
            file = results_file, append = TRUE)
      } else {
        es <- cliff.delta(g1, g2)
        cat(paste("\nCliff's delta for", pair[1], "vs", pair[2], ":\n"), 
            file = results_file, append = TRUE)
      }
      capture.output(print(es), file = results_file, append = TRUE)
    }
  }
}

# Close results file
close(results_file)
```

# Completion and Reporting

```{r completion}
cat("\nAnalysis Complete!\n")
cat("Outputs Generated:\n")
cat("1. Visualization Plots in 'plots/' directory\n")
cat("2. Comprehensive Statistical Report in 'results/statistical_results.txt'\n")
```

# Session Information and Reproducibility

```{r session-info}
# Capture complete environment details
sessionInfo()
```
